Geometric variational problems

### 2020-11-12 NSF2020/Description/specificaim2_ver2.tex
worked on SA1 and small edits elsewhere
edited up to 4.2.2

\subsection{Specific Aim 2: Hydrodynamics}
\label{subsec:specific_aim_2}
In Specific Aim 2, we include hydrodynamics effects by embedding the particles in a solvent. 
As described in \eqref{eq:stokes} and illustrated in Figure \ref{fig:flow_map}, 
the solvent phase is modeled by the Stokes equations for an incompressible, zero Reynolds-number fluid, 
and these equations are coupled to the screened-Laplace equation \eqref{SL} through viscous and hydrophobic stress balance.
Hydrodynamic effects are a relevant detail  
because the rates of biological functions like fusion, fission, and pore dynamics rely on viscous dissipation \cite{RYHAM20112929}. 
%While hydrophobic attraction causes the particles to self-assemble indepedently of the dissipation mechanism, the trajectories and
%equilibrium configurations depend on the fact that the particles travel through a viscous environment. 
Additionally, hydrodynamics interactions are central fabrication of complex microscopic three-dimensional structures \cite{Cho2010}.
Over the past decade, there has been an explosion of interest in small-scale processes that utilize capillary forces dominate,  van der Waals interactions and thermal noise  \cite{Zhang2017}, to coordinated movement and bind  material subunits, \cite{Pandey2011,Leong2007,Reynolds2019, Dasgupta2017,Siontorou2017}.
Finally, coarse-grained and molecular dynamics theories include water either explicitly or implicitly in their models. 
The immediate goal of the present proposal
is to be able to simulate vesicle bilayers in external flows, where the bilayers are represented by a self-assembled collection of rigid particles.

@@

\textbf{Vesicles in background flows.}
Numerically solving the Stokes-screened Laplace system \eqref{SL}, \eqref{eq:stokes} is a nontrivial matter,
and we will develop an efficient, robust, high-order of accuracy in time and space algorithm suited to the problem. 
The challenges are: \textbf{(1)} the equations express a two-way coupling.
This means that the flow changes the shape of the bilayer (or particle suspension, more generally),
and in return the hydrophobic stresses coming from the altered shape imparts a force on the flow. 
\textbf{(2)}
The inputs to the Stokes equations are particle configurations, forces and torques. 
The outputs are the rigid body translation and angular velocties used to update the particle positions (see Figure \ref{fig:flow_map}).
Although the underlying equations are linear, the overall problem is highly nonlinear because the domain is constantly changing.
\textbf{(3)} Self-assembly causes the particles to come into close contact.  
As a result, a exceptional spatial accuracy is required to resolve the fields between adjacent particles.
Finally, the physically relevant elastic properties of bilayer become apparent at large length scales and for large particle-numbers,
which increases the computational complexity of our simulations.

We include a background flow by replacing the third equation in \eqref{eq:stokes} with the condition
$(\mathbf{u} - \mathbf{u}_{\infty})(\mathbf{x}) \to \mathbf{0}$ as $|\mathbf{x}| \to \infty.$ 
The main idea to incorporate the far-field flow is to use the representation 
\begin{equation}
\label{PowerMiranda}
{\bf u} = {\bf u}_{\infty} + K\eta + \sum_{i=1}^N S(\cdot, {\bf a}_i) {\bf F}_i + R(\cdot, {\bf a}_i) {\bf G}_i.
\end{equation}
where $S$ and $R$ are stokeslets and rotlets supported at the respective particle centers 
and $K\eta$ is a layer potential for an unknown density function $\eta.$ 
The representation \eqref{PowerMiranda} automatically satisfies all equations, with the exception of the rigid motion conditions. 
These rigid motion conditions follow by requiring the viscous stress vanishes across the particle boundaries.

We will compare the behavior of our particle-based vesicles in Stokes flow to well-established results.
\todo{What are the known results?}
Researchers have devised models to enforce area incompressibility and
volume conservation in vesicle membrane dynamics. 
\cite{torres-sanchez_millan_arroyo_2019, mahapatra_saintillan_rangamani_2020, Steigmann99, C6SM02452A}.
Our preliminary results show that the particle-based bilayers have the same, large area modulus as that of real lipid bilayer membranes,
and so we expect realistic results from our flow simulations.
Under moderate shear rates, the vesicle elongates and takes on the tank-treading, elliptical shape.
We can directly check for area compressibility conservation properties from the particle simulation,
by tracking the change in area density of the midplane as a function of shear flow rate,  
along with evaluating the tangential divergence of the velocity.

Bilayer membranes have a small permeability to water \cite{323e9a2f0c58487ea82518d7a1f96485},
and so modelers often assume a vesicle membrane that conserves volume. In our particle-based approach, water motion across
the bilayer is limited by the size of the gaps between the particles, which is an artifact of coarse-graining.
Making these gaps smaller comes at the expense of numerical accuracy, and so we will assess if there is a reasonable trade-off
between approximate volume conservation and simulation complexity.

\begin{wrapfigure}[16]{r}{0.30\textwidth}
\centerline{\includegraphics[width=0.30\textwidth]{figures/BIError.pdf}}
\caption{
\label{fig:bierror}  
The false color map shows how numerical quadrature of layer
potentials loses accuracy near the particle boundaries.
The color bar is for $\log_{10}.$ 
}
\end{wrapfigure}

Finally, the particle-based vesicle bilayers have two distinct leaflets,
consisting of the particles in contact with the vesicle interior and 
surrounding fluid respectively. Since there is a nonzero separation between these layers,
one important effect we observe is that the leaflets slide against one another under shear flow. 
This implies that part of the viscous dissipation in the aqueous phase is enhanced by intermonolayer friction
\cite{SHKULIPA2005823, ShkulipaThesis}.
Intermonolayer slip enters zero-thickness membrane models as a velocity jump boundary condition
and leads to multiple solution states \cite{schwalbe_vlahovska_miksis_2010}.
In the present proposal, intermonolayer slip is an artifact of monolayer independence  
and we can control for slip as a function of shear rate, vesicle diameter and particle geometry.

\textbf{Novel reciprocal identities.}
In the course of our study, we have found it useful to devise certain reciprocal identities that
allow us to calculate forces accurately without having to resort to overly complicated numerical integration schemes. 
At issue is that there is a break-down in numerical accuracy for the gradient of water activity near the particle
boundaries (see Figure \ref{fig:bierror}).
This is unfortunate the boundary is exactly where we need the gradients to be accurate in order for the formulas \eqref{forceandtorque}
to be meaningful. To get around this issue, we prove the following identity
  \begin{equation}
    \label{eq:reciprocal}
{\bf F}_{\text{hydro},i} = \sum_{j \neq i} \int_{\partial P_i}[\boldsymbol{\sigma}_{ij} + \boldsymbol{\sigma}_{ji}]\boldsymbol{\nu}\,\dif S,\quad
{\bf G}_{\text{hydro},i} = \sum_{j \neq i} \int_{\partial P_i} ({\bf x}-\mathbf{a}_i) \times [\boldsymbol{\sigma}_{ij} + \boldsymbol{\sigma}_{ji}]\boldsymbol{\nu}\,\dif S
  \end{equation}
for $i = 1\dots, N$. Here $u = \sum_{i=1}^N u_i$ is a sum of solutions $u_i$ the screened Laplace equation,
$\boldsymbol{\sigma}_{ij} = \rho^{-1} u_iu_j {\bf I} + \rho(\nabla u_i \cdot \nabla u_j {\bf I} - 2 \nabla u_i \otimes \nabla u_j)$ and
$[\cdot]$ denotes the jump across $\partial \Omega.$  It turns out that because of \cite{eq:reciprocal},
we no longer require knowledge of the gradient on the boundary, which is enormously beneficial for calculating force and torque.

>>>>

\subsection{Specific Aim 2: Hydrodynamics}
\label{subsec:specific_aim_2}
In Specific Aim 2, we include hydrodynamics effects by embedding the
particles in a solvent.  As described in~\eqref{eq:stokes} and
illustrated in Figure~\ref{fig:flow_map}, the solvent phase is modeled
by the Stokes equations for an incompressible viscous fluid, and these
equations are coupled to the screened-Laplace equation~\eqref{SL}
through viscous and hydrophobic stress balance. Hydrodynamic effects can
not be ignored since the rates of biological functions like fusion,
fission, and pore dynamics rely on viscous
dissipation~\cite{RYHAM20112929}. 
%While hydrophobic attraction causes the particles to self-assemble indepedently of the dissipation mechanism, the trajectories and
%equilibrium configurations depend on the fact that the particles travel through a viscous environment. 
Additionally, hydrodynamic interactions are central for complex
microscopic three-dimensional structures to fabricate~\cite{Cho2010}.
Over the past decade, there has been an explosion of interest in
small-scale processes that utilize capillary forces, van der Waals
interactions, and thermal noise~\cite{Zhang2017}, to coordinate movement
and bind material subunits~\cite{Pandey2011, Leong2007, Reynolds2019,
Dasgupta2017, Siontorou2017}. Finally, coarse-grained and molecular
dynamics theories include water either explicitly or implicitly in their
models. The immediate goal of the present proposal is to simulate
vesicle bilayers in external flows, where the bilayers are represented
by a self-assembled collection of rigid particles.

@@

\subsubsection{Vesicles in background flows}
Numerically solving the Stokes-screened Laplace system~\eqref{SL}
and~\eqref{eq:stokes} is nontrivial, and we will develop efficient,
robust, high-order in time and space algorithms suited to the problem.
The challenges are: {\bf (1)} the equations express a two-way coupling.
This means that the flow changes the shape of the suspended particles,
and in return the geometry-dependent hydrophobic stresses impart a force
on the flow. {\bf (2)} The inputs to the Stokes equations are particle
configurations, forces, and torques. The outputs are the rigid body
translation and angular velocities used to update the particle positions
(see Figure~\ref{fig:flow_map}). Although the underlying equations are
linear, the overall problem is highly nonlinear because the domain is
constantly changing. {\bf (3)} Self-assembly causes the particles to
come into close contact. As a result, an exceptional spatial accuracy is
required to resolve the fields between adjacent particles. Finally, the
physically relevant elastic properties of bilayer become apparent at
large length scales and for large particle-numbers, which increases the
computational complexity of our simulations.

We include a background flow by replacing the third equation in
\eqref{eq:stokes} with the condition $(\mathbf{u} -
\mathbf{u}_{\infty})(\mathbf{x}) \to \mathbf{0}$ as $|\mathbf{x}| \to
\infty$. To incorporate the far-field flow, we use the representation 
\begin{align}
\label{PowerMiranda}
  {\bf u} = {\bf u}_{\infty} + K\boldsymbol{\eta} + 
    \sum_{i=1}^N S(\cdot, {\bf a}_i) {\bf F}_i + 
                 R(\cdot, {\bf a}_i) {\bf G}_i.
\end{align}
where $S$ and $R$ are stokeslets and rotlets supported at the respective
particle centers and $K\boldsymbol{eta}$ is a layer potential for the
unknown density function $\boldsymbol{\eta}.$ The
representation~\eqref{PowerMiranda} automatically satisfies all
equations, with the exception of the rigid motion conditions. The rigid
motion conditions follow by requiring the viscous stress vanishes across
the particle boundaries.

We will compare the behavior of our particle-based vesicles in Stokes
flow to well-established results. \todo{What are the known results?}
Researchers have devised models to enforce area incompressibility and
volume conservation in vesicle membrane
dynamics~\cite{torres-sanchez_millan_arroyo_2019,
mahapatra_saintillan_rangamani_2020, Steigmann99, C6SM02452A}. Our
preliminary results show that the particle-based bilayers have the same
large area modulus as real lipid bilayer membranes, so we expect
realistic results from our flow simulations. Under moderate shear rates,
the vesicle elongates and takes on the tank-treading  elliptical shape.
We can directly check for area compressibility conservation properties
from the particle simulation by tracking the change in area density of
the midplane as a function of shear flow rate, along with evaluating the
tangential divergence of the velocity.

Bilayer membranes have a small permeability to water
\cite{323e9a2f0c58487ea82518d7a1f96485}, and modelers often assume a
vesicle membrane that conserves volume. In our particle-based approach,
water motion across the bilayer is limited by the size of the gaps
between the particles, which is an artifact of coarse-graining.  Making
these gaps smaller comes at the expense of numerical accuracy, and we
will assess if there is a reasonable trade-off between approximate
volume conservation and simulation complexity.

\begin{wrapfigure}[13]{r}{0.30\textwidth}
\centerline{\includegraphics[width=0.30\textwidth]{figures/BIError.pdf}}
  \vspace{-8pt}
\caption{
\label{fig:bierror}  
\footnotesize The false color map shows how numerical quadrature of
  layer potentials loses accuracy near the particle boundaries.  The
  color bar is for $\log_{10}.$}
\end{wrapfigure}

Finally, the particle-based vesicle bilayers have two distinct leaflets
consisting of the particles in contact with the vesicle interior, and
those in contact with the surrounding fluid. Since there is a nonzero
separation between these layers, an important effect we observe is that
the leaflets slide against one another under shear flow.  This implies
that part of the viscous dissipation in the aqueous phase is enhanced by
intermonolayer friction~\cite{SHKULIPA2005823, ShkulipaThesis}.
Intermonolayer slip enters zero-thickness membrane models as a velocity
jump boundary condition and leads to multiple solution
states~\cite{schwalbe_vlahovska_miksis_2010}.  In the present proposal,
intermonolayer slip is an artifact of monolayer independence  and we can
control for slip as a function of shear rate, vesicle diameter, and
particle geometry.

\subsubsection{Novel reciprocal identities}
In the course of our study, we have found it useful to devise certain reciprocal identities that
allow us to calculate forces accurately without having to resort to overly complicated numerical integration schemes. 
At issue is that there is a break-down in numerical accuracy for the gradient of water activity near the particle
boundaries (see Figure \ref{fig:bierror}).
This is unfortunate the boundary is exactly where we need the gradients to be accurate in order for the formulas \eqref{forceandtorque}
to be meaningful. To get around this issue, we prove the following identity
  \begin{equation}
    \label{eq:reciprocal}
{\bf F}_{\text{hydro},i} = \sum_{j \neq i} \int_{\partial P_i}[\boldsymbol{\sigma}_{ij} + \boldsymbol{\sigma}_{ji}]\boldsymbol{\nu}\,\dif S,\quad
{\bf G}_{\text{hydro},i} = \sum_{j \neq i} \int_{\partial P_i} ({\bf x}-\mathbf{a}_i) \times [\boldsymbol{\sigma}_{ij} + \boldsymbol{\sigma}_{ji}]\boldsymbol{\nu}\,\dif S
  \end{equation}
for $i = 1\dots, N$. Here $u = \sum_{i=1}^N u_i$ is a sum of solutions $u_i$ the screened Laplace equation,
$\boldsymbol{\sigma}_{ij} = \rho^{-1} u_iu_j {\bf I} + \rho(\nabla u_i \cdot \nabla u_j {\bf I} - 2 \nabla u_i \otimes \nabla u_j)$ and
$[\cdot]$ denotes the jump across $\partial \Omega.$  It turns out that because of \cite{eq:reciprocal},
we no longer require knowledge of the gradient on the boundary, which is enormously beneficial for calculating force and torque.
